The Impact of Anthropogenic Forcing on ENSO Amplitude

### 2022-01-24 presentations/presentation/notes.tex
stuff lol

% Created 2021-10-31 Sun 14:05
% Intended LaTeX compiler: pdflatex
\documentclass[little]{basic}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\setcounter{secnumdepth}{1}
\author{Benjamin Goldman}
\date{\today}
\title{Presentation Notes}
\hypersetup{
 pdfauthor={Benjamin Goldman},
 pdftitle={Presentation Notes},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 27.2 (Org mode 9.5)}, 
 pdflang={English}}
\begin{document}

@@

\section{General:}
\label{sec:orgf4efe11}
\begin{itemize}
\item Remember to say meaning of axes and colors
\item Talk slower
\item Less detail in results, more in intro?
\end{itemize}

@@

\label{sec:org197abb8}

@@

\label{sec:org7e341be}

@@

\label{sec:orgeb7a331}

@@

\label{sec:org08234e7}

@@

\label{sec:org0ec8ade}

@@

\label{sec:org902ab02}

@@

\label{sec:orgf798249}

@@

\label{sec:org177c1d6}

@@

\label{sec:org1ab33ee}

@@

\label{sec:org84e14f6}

@@

\label{sec:org86e31d8}

@@

\label{sec:org2e78332}

@@

\label{sec:org0f6ba67}

@@

\label{sec:orge16e5c1}

@@

\label{sec:orgb07762a}

@@

\label{sec:org2b52160}

@@

\label{sec:orgbdaf9d4}

@@

\label{sec:orga32dcde}

@@

\section{Conclusion}
\label{sec:org46c3af5}
\subsection*{Conclusions and Discussion}
\label{sec:org10adbad}
\subsection*{Application, Limitation, and Next Steps}
\label{sec:org10fbd6a}
\subsection*{Acknowledgments}
\label{sec:orgf4b1b34}
\item Thank you to NCAR for running the climate models
\item Thank you to my teacher, family, and mentor
\item Thank you to the software makers
\subsection*{References}
\label{sec:orgd9329c4}
\end{document}

>>>>

% Created 2022-01-18 Tue 20:19
% Intended LaTeX compiler: pdflatex
\documentclass[little]{basic}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\setcounter{secnumdepth}{1}
\author{Benjamin Goldman}
\date{\today}
\title{Presentation Notes}
\hypersetup{
 pdfauthor={Benjamin Goldman},
 pdftitle={Presentation Notes},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 27.2 (Org mode 9.5)}, 
 pdflang={English}}
\begin{document}

@@

\section{General:}
\label{sec:org2299927}
\begin{itemize}
\item Remember to say meaning of axes and colors
\item Talk slower
\item Less detail in results, more in intro?
\end{itemize}

@@

\label{sec:org95ba26f}

@@

\label{sec:orgee84b8b}

@@

\label{sec:org68534c1}

@@

\label{sec:orga3b6ea1}

@@

\label{sec:org3e6158d}

@@

\label{sec:orgc27209d}

@@

\label{sec:org4631399}

@@

\label{sec:org64f4a3a}

@@

\label{sec:org41ce2a5}

@@

\label{sec:org4f1184c}

@@

\label{sec:orgea6d725}

@@

\label{sec:org996c977}

@@

\label{sec:org8ffde0b}

@@

\label{sec:org8f32f98}

@@

\label{sec:org5844385}

@@

\label{sec:org80018e5}

@@

\label{sec:org403aa1c}

@@

\label{sec:orgb0f8166}

@@

I conducted a wavelet analysis to identify changes to specific frequency bands of ENSO. One way of thinking about wavelet analysis is using a radio to listen to a variety of frequencies, all at once, except with data rather than radio waves. So figure 7 shows, the power, similar to loudness, of the ENSO signal where the period is greater than three years and less than three years, for both the CESM1 and CESM2. Looking at CESM1, much of the increase in power over the 20th century is in the longer-period band, meaning that much of the increase in ENSO intensity is happening in the lower frequencies. Looking at CESM2, one interesting feature is that the decreasing strength we saw earlier is mostly happening in the longer frequency. These results suggest that longer frequency ENSO is more succeptible to climate change than ENSO with a faster cycle.

\section{Conclusion}
\label{sec:org92a7dae}
\subsection*{Conclusions and Discussion}
\label{sec:orge52295e}
\begin{itemize}
\item Slide is notes
\end{itemize}
\subsection*{Application, Limitation, and Next Steps}
\label{sec:org0c817e6}
\begin{itemize}
\item Slide is notes
\end{itemize}
\subsection*{Acknowledgments}
\label{sec:orgeaf66eb}
\begin{itemize}
\item Thank you to NCAR for running the climate models
\item Thank you to my teacher, family, and mentor
\item Thank you to the software makers
\end{itemize}
\subsection*{References}
\label{sec:org0630c81}
\end{document}

### 2022-01-24 presentations/presentation/presentation.tex
stuff lol

\begin{frame}{Climate Change and Variability}
  \begin{columns}
    \column{0.4\textwidth}
    \begin{itemize}
    \item Global warming
    \item Long-term trends vs short-term randomness
    \end{itemize}
    \column{0.6\textwidth}
    \myfig{intro_fig_3.pdf}{Global mean land air temperature in GISSTEMP 4 dataset. \citep{gistemp2019giss} and \citep{lenssen2019improvements}}{woohoo}
  \end{columns}
\end{frame}

\begin{frame}{Climate forcing}
  \begin{columns}
    \column{0.5\textwidth}
    \begin{itemize}
    \item \alert{Forcing}: any external factor that affects climate.
      \begin{description}
      \item[\alert{GHG}] Greenhouse gasses
      \item[\alert{AER}] Aerosols (natural: volcanic ash, artificial: smoke)
      \item[\alert{BMB}] Biomass burning
      \item[\alert{LULC}] Land use/cover (deforestation, desertification)
      \end{description}
    \end{itemize}
    \column{0.6\textwidth}
    \myfig{greenhouse_Effect.jpg}{Factors that contribute to the greenhouse effect. \url{https://www.coolaustralia.org/the-greenhouse-effect-secondary}}{this}
  \end{columns}
\end{frame}

\begin{frame}{El Niño (ENSO)}
  \begin{figure}
    \begin{columns}
      \column{0.5\textwidth}
      \begin{itemize}
      \item ENSO = El Niño/Southern Oscillation
      \item Warming and cooling of the Pacific Ocean.
      \item Affects human societies through temperature and rainfall. \citep{ropelewski1987global}
      \item May be affected by climate change.
      \end{itemize}
      \caption{Comparison of SST anomaly between 1975 La Niña event and 1997 El Niño event in HadISST 1 dataset. \citep{rayner2003global}}
      \column{0.5\textwidth}
      \includegraphics[width = \textwidth]{figures/intro_fig.pdf}
    \end{columns}
  \end{figure}
\end{frame}

\begin{frame}{Review of Literature}
  \begin{itemize}
  \item ENSO's properties observed vary across different decades. \citep{lubbecke2014assessing}.
  \item ENSO responds to external forcing.
  \item Weakened ENSO during the Ice Age due to reduced CO$_2$ levels \citep{zhu2017reduced}.
  \item Models show possible increasing ENSO activity in the future \citep{zheng2017response} and \citep{maher2018enso}.
  \end{itemize}
\end{frame}

\begin{frame}{Gap and Questions}
  \alert{Gap}
  \begin{itemize}
  \item Little research using a large ensemble to examine the effect of individual factors on ENSO.
  \item Considerable disagreement between studies on whether ENSO will strengthen or weaken due to global warming
  \end{itemize}
  \alert{Questions}
  \begin{description}
  \item[\alert{What?}] Do the CESM1 and CESM2 predict increased or decreased ENSO intensity in the future?
  \item[\alert{Why?}] Is the predicted increase (or decrease) due to human activities?
  \item[\alert{How?}] What processes are causing greenhouse gasses and aerosols to affect ENSO?
  \end{description}
\end{frame}

@@

\begin{frame}{Model Setup (Data)}
  \begin{itemize}
  \item Community Earth System Model (CESM) Versions 1 and 2 \citep{kay2015community} \citep{danabasoglu2020community}.
  \item Predicts climate over 21st century with global warming.
  \item 40-50 simulations per ensemble.
  \item Control simulation with pre-1850 forcing levels.
  \item Single forcing ensembles that represent influence of single factor.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Correlation With Ocean Temperature}
  \begin{columns}
    \column{0.65\textwidth}
    \begin{itemize}
    \item Correlation coefficient between ocean temperature and ENSO amplitude.
    \item Negative coefficient in subsurface layer.
    \item Positive coefficient in surface layer.
    \item Suggests that ocean stratification may be mediating global warming influence on ENSO.
    \item Difference in heating modifies mechanics of ENSO cycle.
    \end{itemize}
    \column{0.5\textwidth}
    \myfig{tempdt.pdf}{Correlation coefficient between ENSO amplitude and ocean temperature in equatorial cross-section in the fully-forced CESM1 ensemble}{tempdt}
  \end{columns}
\end{frame}

@@

\begin{frame}{Discussion}
  \begin{itemize}
  \item Rising greenhouse gas levels increase Pacific Ocean stratification, strengthening ENSO cycle.
  \item Aerosol influence is nonlinear because aerosol levels are not purely increasing.
  \item Stronger ENSO may lead to greater temperature variability and extreme weather.
  \item CESM1 and CESM2 conflict in their prediction of the changes to ENSO's frequency.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Role of Mentor and Student}
  \begin{columns}[t]
    \column{.4\textwidth}
    Student:
    \begin{itemize}
    \item Analyze raw data on computer
    \item Produce graphics for analysis and publication
    \item Write documentation
    \item Identify key features of results
    \end{itemize}
    \column{.4\textwidth}
    Mentor:
    \begin{itemize}
    \item Review student writing
    \item Interpret results in the context of climatology
    \item Conduct parallel analysis
    \item Provide raw data from facility
    \end{itemize}
  \end{columns}
\end{frame}

>>>>

\begin{frame}{Climate Change: Global Warming vs. Climate Variability}
  \begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/intro_fig_3.pdf}
    \caption{Global mean land air temperature in GISSTEMP 4 dataset. \citep{gistemp2019giss} and \citep{lenssen2019improvements}}
    \label{fig:woohoo}
  \end{figure}
\end{frame}

\begin{frame}{Greenhouse Gasses and Friends}
  \begin{columns}
    \column{0.5\textwidth}
    \begin{itemize}
    \item \alert{Forcing}: any external factor that affects climate.
      \begin{description}
      \item[\alert{GHG}] Greenhouse gasses
      \item[\alert{AER}] Aerosols (natural: volcanic ash, artificial: smoke)
      \item[\alert{BMB}] Biomass burning
      \item[\alert{LULC}] Land use/cover (deforestation, desertification)
      \end{description}
    \end{itemize}
    \column{0.6\textwidth}
    \myfig{greenhouse_Effect.jpg}{Factors that contribute to the greenhouse effect. \url{https://www.coolaustralia.org/the-greenhouse-effect-secondary}}{this}
  \end{columns}
\end{frame}

\begin{frame}{El Niño, La Niña, and ENSO}
  \begin{figure}
    \begin{columns}
      \column{0.5\textwidth}
      \begin{itemize}
      \item ENSO = El Niño/Southern Oscillation
      \item El Niño is the warm phase, and La Niña is the cool phase.
      \item Warming and cooling of the Pacific Ocean.
      \item Affects human societies through temperature and rainfall. \citep{ropelewski1987global}
      \end{itemize}
      \caption{Comparison of SST anomaly between 1975 La Niña event and 1997 El Niño event in HadISST 1 dataset. \citep{rayner2003global}}
      \column{0.6\textwidth}
      \includegraphics[width = \textwidth]{figures/intro_fig.pdf}
    \end{columns}
  \end{figure}
\end{frame}

\begin{frame}{Review of Literature}
  \begin{itemize}
  \item ENSO's properties observed vary across different decades. \citep{lubbecke2014assessing}.
  \item Weakened ENSO during the Ice Age due to reduced CO$_2$ levels \citep{zhu2017reduced}.
  \item Models show possible increasing ENSO activity in the future \citep{zheng2017response} and \citep{maher2018enso}.
  \end{itemize}
\end{frame}

\begin{frame}{Gap and Questions}
  \alert{Gap}
  \begin{itemize}
  \item Little research using a large ensemble to examine the effect of individual factors on ENSO.
  \item Considerable disagreement between studies on whether ENSO will strengthen or weaken due to global warming
  \end{itemize}
  \alert{Questions}
  \begin{enumerate}
  \item Do the CESM1 and CESM2 predict increased or decreased ENSO intensity in the future?
  \item Is the predicted increase (or decrease) due to human activities?
  \end{enumerate}
\end{frame}

@@

\begin{frame}{Data: Community Earth System Model}
  \begin{itemize}
  \item Community Earth System Model (CESM) Versions 1 and 2 \citep{kay2015community} \citep{danabasoglu2020community}.
  \item Predicts climate over 21st century with global warming.
  \item Ensemble: collection of multiple simulations.
  \item Single forcing ensembles that represent influence of single factor.
  \end{itemize}
\end{frame}

@@

% \begin{frame}{Correlation With Ocean Temperature}
%   \begin{columns}
%     \column{0.65\textwidth}
%     \begin{itemize}
%     \item Correlation coefficient between ocean temperature and ENSO amplitude.
%     \item Negative coefficient in subsurface layer.
%     \item Positive coefficient in surface layer.
%     \item Suggests that ocean stratification may be mediating global warming influence on ENSO.
%     \item Difference in heating modifies mechanics of ENSO cycle.
%     \end{itemize}
%     \column{0.5\textwidth}
%     \myfig{tempdt.pdf}{Correlation coefficient between ENSO amplitude and ocean temperature in equatorial cross-section in the fully-forced CESM1 ensemble}{tempdt}
%   \end{columns}
% \end{frame}

@@

\begin{frame}{Discussion}
  \begin{itemize}
  \item Rising greenhouse gas levels strengthen ENSO cycle.
  \item Aerosol influence is nonlinear because aerosol levels are not purely increasing.
  \item Stronger ENSO may lead to greater temperature variability and extreme weather.
  \item External forcing affects lower frequency ENSO more.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Role of Mentor and Student}
  \begin{columns}[t]
    \column{.5\textwidth}
    Student:
    \begin{itemize}
    \item Analyze raw data on computer
    \item Produce graphics for analysis and publication
    \item Write documentation
    \item Identify key features of results
    \end{itemize}
    \column{.5\textwidth}
    Mentor:
    \begin{itemize}
    \item Review student writing
    \item Interpret results in the context of climatology
    \item Conduct parallel analysis
    \item Provide raw data from facility
    \end{itemize}
  \end{columns}
\end{frame}
